%=============================================================================
\section{Discussion}
\label{sec:discussion}
%=============================================================================

\subsection{Loop collapse as a structural correlate of IBD dysbiosis}

The \ibd\ finding requires careful interpretation at two levels.
At the group level, the bootstrap procedure detects consistent topological
differences between IBD and healthy co-occurrence networks: all six $H_1$
features reach the permutation floor ($p = 0.002$) at every taxon-set
size, and results are preserved after covariate matching.
This establishes that the \emph{population-level network architecture}
differs between groups.

At the individual level, however, per-sample topology yields modest
effect sizes ($|d| = 0.04$--$0.37$), comparable to Shannon diversity
($|d| = 0.44$).
The large Cohen's $d$ values reported in Table~\ref{tab:ibd_results}
($|d| = 1.2$--$2.3$) reflect the separation of bootstrap-iteration-level
summaries, each aggregating 100 samples; they are not comparable to
standard per-sample benchmarks.
This distinction is methodologically important: the bootstrap $d$ confirms
that the topological gap is detectable at virtually any randomly drawn
subsample, but it does not imply that individual patients can be classified
with high accuracy from topology alone.
The per-sample validation (Section~\ref{subsec:persample_validation})
demonstrates that topology's practical value lies in providing complementary
information to Shannon diversity (combined AUC $= 0.674$ vs.\ $0.645$
Shannon-only), not in replacing it.

The direction of the finding is consistent with known IBD compositional
pathology.
Butyrate-producing Firmicutes (\textit{Faecalibacterium},
\textit{Roseburia}, \textit{Lachnospiraceae}) occupy dense, multi-partner
co-occurrence neighborhoods in healthy gut networks; their depletion in
\ibd\ would be expected to dissolve the loops anchored around
them~\citep{gevers2014treatment,halfvarson2017dynamics}.
Whether these particular taxa drive the loop collapse, or whether the
entire gram-positive anaerobe compartment contracts uniformly, is directly
addressed by the \loo\ attribution in Section~\ref{subsec:attribution}:
\textit{Oscillospira} and \textit{Blautia} emerge as the dominant
differential anchors in the AGP 16S data.

A more mechanistic reading is that the healthy gut co-occurrence graph
approximates a modular, redundant architecture: butyrate producers and
fiber-degrading specialists form mutually reinforcing exchange loops in
which each partner both consumes the metabolic byproducts of others and
releases substrates that sustain them.
Persistent loops in the $H_1$ sense require at least three taxa forming
a correlation triangle strong enough to survive across a wide filtration
range; the co-occurrence architecture of butyrate producers plausibly
satisfies this criterion.
When these taxa are depleted or become numerically erratic under
\ibd-associated inflammation, the triangles dissolve before they
consolidate, producing a persistence diagram with lower total mass and
shorter feature lifetimes.
What remains is not necessarily a sparser network in the edge-count
sense---Proteobacterial blooms can increase certain pairwise correlations
in \ibd~\citep{halfvarson2017dynamics}---but a topologically impoverished
one: shorter-lived, less reinforced cycles.
This interpretation predicts that the Proteobacterial compartment, while
elevated in \ibd, does not substitute structurally for the anaerobic
Firmicutes it displaces---a prediction consistent with the \loo\ attribution
finding that \textit{Clostridiaceae} and \textit{Lachnospiraceae}, not any
Proteobacterial genus, are the significant healthy anchors in AGP
(Section~\ref{subsec:attribution}).

\subsection{Structural reorganization versus abundance enrichment in diet}

The bifurcated diet signal (fewer total loops, longer-lived individual
loops for plant-based eaters) resists a simple narrative.
It is not that plant-based microbiomes are topologically richer in an
unqualified sense; they appear to support a different \emph{architecture}:
a sparser set of co-occurrence partnerships, but ones with more robust,
threshold-spanning correlations.
One mechanistic hypothesis consistent with this pattern is dietary
fibre-driven selection: high-fibre diets enforce strong metabolic
dependencies among a constrained set of fibre-degrading consortia
(principally \textit{Prevotella}, \textit{Ruminococcus}, and
\textit{Bifidobacterium} genera), producing persistent loops, while
reducing the proliferation of opportunistic taxa whose co-occurrences
are numerous but shallow~\citep{louis2017formation}.
This is speculative; validating it would require superimposing fibre
intake quantification on loop attribution.

The partial attenuation of the diet signal after matching (3/6 features
surviving FDR) counsels against strong claims.
It is plausible that a fraction of the full-group diet effect is
sociological, given that plant-based eaters in \agp\ are skewed toward higher
education, younger age, and metropolitan residence, rather than purely
physiological in origin.
The three features that do survive matching (mean lifetime, max
lifetime, max Betti-1) are precisely the ones reflecting loop durability,
which are less susceptible to demographic covariates than loop count.

\subsection{The antibiotic null}

That a 6-month antibiotic history window produces no coherent topological
signal after matching is not surprising, and should not be over-interpreted
as evidence that antibiotics are topologically inert.
The exposure category aggregates single-course amoxicillin prescriptions,
prolonged prophylactic macrolide regimens, and fluoroquinolone treatment:
perturbations that differ by orders of magnitude in their ecological
destructiveness~\citep{petersen2014defining}.
A better-powered analysis would stratify by antibiotic class and quantify
time since last exposure as a continuous variable.

\subsection{Relation to prior topological analyses of microbiome data}

The deployment of persistent homology in microbiology is recent and sparse.
C\'{a}mara~\citep{camara2017topological} surveyed the emerging application
of $H_0$ and $H_1$ features to genomic and microbial data, identifying
trajectory-level topology---tracking how network structure evolves across
dense temporal sampling---as a particularly promising regime for small-$n$,
longitudinal designs.
Our design occupies the complementary regime: large cross-sectional $n$,
sparse metadata, no temporal structure, but high statistical power for
detecting mean differences between clinically defined groups.
The two approaches are complementary rather than competing.

More broadly, TDA has been applied to protein contact networks to predict
folding stability~\citep{wasserman2018topological}, to neuroimaging data to
map functional connectivity~\citep{bubenik2015statistical}, and to materials
microstructure to predict fracture~\citep{ghrist2008barcodes}.
In each domain, the practical justification for persistent homology over
graph-theoretic alternatives is the same as here: the features of interest
are not expressible as functions of individual edges, but emerge from the
collective geometry of multi-node motifs that thresholding-based networks
miss or misrepresent.

To test this claim directly, we ran a parallel benchmark in which the paired
bootstrap framework was applied to standard graph-theoretic metrics (edge density,
clustering coefficient, transitivity, modularity, number of connected components)
at three absolute-correlation thresholds ($|r| \geq 0.2, 0.3, 0.4$) in the
IBDMDB cohort.
Standard metrics do detect the IBD signal---transitivity achieves $|d| = 1.20$
at threshold 0.2, and modularity reaches $|d| = 0.91$ at 0.3---but their
performance is \emph{threshold-dependent}: edge density is significant at
$|r| \geq 0.2$ and $0.4$ but not $0.3$; clustering coefficient is significant
at $0.2$ and $0.3$ but not $0.4$.
By contrast, all six TDA features are significant at the permutation floor
($p = 0.002$) without requiring any threshold choice, because persistent
homology integrates across the full filtration.
The highest single network-metric effect size ($|d| = 2.45$ for number of
connected components at $|r| \geq 0.4$) exceeds the best TDA feature
($|d| = 1.46$ for total persistence), but this metric is essentially $H_0$
(zeroth homology), which TDA also computes; the distinctive contribution of
persistent homology lies in the $H_1$ loop features, which have no
graph-theoretic equivalent.
Moreover, the $H_2$ stress test (Section~\ref{subsec:ibdmdb}) reveals
that topological simplification in dysbiosis is not confined to
one-cycles: two-dimensional voids are equally depleted in IBD networks
($|d| = 1.13$--$2.28$, all $p < 0.005$), suggesting that the entire
simplicial complex---not just its loop structure---collapses under
dysbiosis.
This multi-dimensional topological degradation is a novel finding with no
analogue in standard network-metric analyses.

The group-level topological signal is consistent with this general principle:
persistent homology aggregates information from many pairwise correlations
simultaneously, detecting network-wide reorganization that is difficult to
capture with single-sample diversity indices.
However, per-sample validation (Section~\ref{subsec:persample_validation})
reveals that this advantage is modest at the individual patient level,
with topology providing complementary rather than dominant predictive power.
The IBDMDB replication (Section~\ref{subsec:ibdmdb}) extends these results
to a second dataset and shows that topological features are sensitive not
only to the categorical disease label but to continuous biomarker severity,
a property that prior applications of TDA in biology have not demonstrated
in a clinical context.

\subsection{Limitations}
\label{subsec:limitations}

\paragraph{Bootstrap effect size interpretation.}
The Cohen's $d$ values from the bootstrap procedure
(Tables~\ref{tab:ibd_results},~\ref{tab:ibdmdb_ibd}) measure
separation between distributions of group-level network statistics, not
between individual patient measurements.
Each bootstrap iteration produces one scalar per group (summarizing 100
samples), so the variance of these iteration-level values is substantially
smaller than the variance of per-patient measurements, inflating $d$ relative
to conventional benchmarks.
Per-sample validation (Section~\ref{subsec:persample_validation}) confirms
that individual-level effects are small ($|d| \leq 0.37$).
The bootstrap $d$ remains valid for its intended purpose---quantifying the
reliability of the group-level separation---but should not be interpreted
using standard per-sample thresholds.

\paragraph{Cross-sectional topology.}
All \agp\ records are single-timepoint observations.
The causal directionality (whether topological simplification precedes \ibd\
onset, coincides with active inflammation, or lingers through clinical
remission) cannot be resolved without longitudinal sampling.

\paragraph{ASV-level resolution.}
The pipeline operates on Deblur ASVs without genus-level aggregation.
Strain-level co-occurrence heterogeneity within named genera may both inflate
and obscure loop structure, and interoperability with genus-level literature
is impaired.

\paragraph{Diversity-metric benchmark (addressed).}
Section~\ref{subsec:classification} extends the original Shannon benchmark to
include Aitchison $\beta$-diversity (Euclidean distance in CLR space;
phylogeny-free analog of weighted UniFrac) and Bray--Curtis dissimilarity,
both embedded via nested PCoA with no test-fold leakage.
Aitchison-PCoA(10) achieves AUC $= 0.734$ (LR), outperforming Shannon
($0.688$) and topology scalars ($0.668$), and the six H$_1$ scalars do not
improve over Aitchison when combined.
The remaining benchmark gap is whether richer topological representations
(persistence images, kernel embeddings) can close the AUC gap to
Aitchison-PCoA; that question is deferred to future work.

\paragraph{Transductive leakage (addressed).}
Section~\ref{subsec:inductive} reports an inductive benchmark in which
co-occurrence networks are computed separately on training and test
partitions, eliminating the shared-correlation-matrix leakage present in
the transductive design.
The transductive--inductive gap is $\Delta\text{AUC} \approx 0.013$ for
topology alone, small relative to cross-seed variance ($\text{SD} \approx 0.03$).
A 1{,}000-shuffle label-permutation test confirms $p < 0.001$, and
systematic robustness checks (20 seeds, 4 split ratios, 4 CV fold counts,
4 subsampling fractions) show that the classification signal is stable and
not an artefact of any particular evaluation configuration.
Topology features do, however, require larger sample sizes than Shannon to
estimate reliably: at 25\% subsampling, topology AUC degrades by $0.027$
while Shannon remains unchanged.

\paragraph{Single cohort (addressed in part).}
All primary \agp\ results originate from a self-recruited citizen-science
population with attendant ascertainment biases.
Section~\ref{subsec:ibdmdb} reports an independent replication in the
IBDMDB clinical cohort~\citep{lloydprice2019multi}, which uses a different
sequencing platform, bioinformatics pipeline, and population (recruited IBD
patients with endoscopic confirmation rather than self-reported community
participants).
The replication is qualitatively successful for the \ibd\ finding.
Residual concerns include the possibility that both datasets share
confounders related to medication (immunosuppressants, biologics) that
are not controlled in either analysis.

\paragraph{Loop attribution completed; biological validation pending.}
Section~\ref{subsec:attribution} reports taxon-level \loo\ attribution in
both datasets.
The aggregate statistical claim (IBD has topologically simpler networks) is
now paired with a taxon-resolved candidate list: \textit{Oscillospira},
\textit{Blautia}, \textit{Lachnospiraceae}, and \textit{Clostridiaceae} in
the AGP 16S data; \textit{Bacteroides finegoldii}, \textit{B.\ thetaiotaomicron},
\textit{Flavonifractor plautii}, and \textit{Clostridium nexile} in the
IBDMDB WGS data.
The remaining gap is biological validation: whether these taxa truly form
the cross-feeding exchange loops predicted by their composite scores, or
whether their high scores reflect shared correlation partners that happen to
track disease status, requires targeted metabolomic or in vitro co-culture
experiments.

\paragraph{Inflammatory biomarker data (partially addressed).}
The \agp\ dataset contains no fecal calprotectin or other inflammatory
biomarkers.
The IBDMDB replication (Section~\ref{subsec:ibdmdb}) partially addresses
this: calprotectin-stratified comparisons confirm that topological
contraction in loop count tracks mucosal inflammation, while revealing
a nuanced secondary pattern (longer-lived individual loops at high
calprotectin) that warrants mechanistic investigation.
Whether loop collapse reflects impaired butyrate-mediated Treg
maintenance remains unresolved in either cohort.

\paragraph{Bootstrap non-independence.}
The 200 paired iterations are constructed from overlapping samples (a sample
from the IBD group can appear in multiple iterations), so the deltas
$\delta^{(1)}, \ldots, \delta^{(200)}$ are not strictly independent.
For the IBD comparison, where the smaller group has $n_A = 165$
and each iteration draws 100 without replacement, any two iterations
share on average $100^2 / 165 \approx 61$ IBD samples; a conservative
effective sample size for IBD-group-driven variation is therefore
$n_\text{eff} \approx 165$ rather than 200.
The healthy-group overlap is negligible by comparison
($100^2 / 3{,}084 \approx 3$ shared samples per pair).
The sign-flip permutation operates on the same correlated set of deltas
and thereby conserves this correlation structure under the null, which is
appropriate.
However, the Wilcoxon signed-rank test assumes independence across the
200 deltas; applying it to correlated observations inflates the test
statistic and underestimates the $p$-value.
This is one reason Wilcoxon results are treated as confirmatory only and
never enter any FDR calculation.
