%=============================================================================
\section{Conclusion}
\label{sec:conclusion}
%=============================================================================

We applied persistent homology to microbial co-occurrence networks derived
from 3,409 fecal samples in the American Gut Project and independently
validated the core finding in the IBDMDB multi-omics cohort.
Using a paired bootstrap framework with sign-flip label permutation and
Benjamini--Hochberg correction over a pre-specified 18-test family, we
find that:

\begin{enumerate}[leftmargin=*, itemsep=3pt]

  \item \textbf{\ibd\ (UC + Crohn's) is associated with consistent
  attenuation of co-occurrence loop structure across all six features.}
  Effect sizes of $d = 0.75$--$2.38$, a null permutation floor at every
  feature, and complete preservation after age/sex/BMI covariate matching
  make this the most robust finding in the study.
  It replicates at taxon-set sizes 50, 80, and 120 within \agp, and is
  reproduced in IBDMDB WGS data using a fully independent pipeline.

  \item \textbf{The loop-collapse signal tracks a continuous inflammation
  biomarker.}
  Fecal calprotectin correlates with topological features in IBDMDB,
  grounding the binary IBD label in a quantitative measure of mucosal
  inflammation and strengthening the mechanistic interpretation.

  \item \textbf{Plant-based diet restructures rather than uniformly
  enriches co-occurrence topology.}
  Fewer but longer-lived loops, partially robust to demographic
  matching (3/6 features FDR-significant post-match), suggest a
  fibre-driven structural re-partitioning of the microbial interaction
  network.

  \item \textbf{Antibiotic recency, as defined in \agp, provides no
  coherent topological signal.}
  The comparison is underpowered for topological inference given the
  heterogeneous exposure window and absence of class stratification.

  \item \textbf{Leave-one-out attribution identifies taxon-level anchors.}
  \textit{Oscillospira} and \textit{Blautia} are the dominant healthy loop
  anchors in the AGP 16S dataset; \textit{Bacteroides}-clade species
  dominate the IBDMDB WGS ranking.
  Seven genera appear in both top-20 lists, providing named biological
  candidates for the loop collapse.

  \item \textbf{An extended classification benchmark clarifies topology's role.}
  Aitchison $\beta$-diversity (10 PCoA axes, no test-fold leakage) is the
  strongest single predictor (AUC $= 0.734$, logistic regression), exceeding
  Shannon ($0.688$) and topology scalars alone ($0.668$).
  Topology scalars do not improve over Aitchison when combined, indicating that
  the six H$_1$ scalars are largely subsumed by the full Aitchison distance
  geometry in prediction.
  The mechanistic value of topology---loop attribution, biomarker correlation,
  structural interpretability---is distinct from and not captured by AUC
  comparisons alone.

\end{enumerate}

The paired bootstrap design, pre-fixed global taxon selection, and
leave-one-out topological attribution pipeline introduced here are reusable
components for any group-level microbiome TDA application.
All code, data retrieval scripts, result tables, and figure generation
pipelines are publicly available at
\href{https://github.com/adam-jfkhs/Microbiome-TDA}%
{\texttt{github.com/adam-jfkhs/Microbiome-TDA}}.

The most pressing open extensions of this work are:
(i) longitudinal tracking of topological features within individual patients
across disease flares and remission, to test whether loop complexity is a
dynamic marker of disease activity rather than a static cross-sectional
trait;
(ii) richer topological representations---persistence images, kernel
embeddings, or $H_2$ cavity counts---which are expected to widen the
topology-versus-Shannon AUC gap beyond the current $0.02$ margin;
and (iii) integration with metatranscriptomic and metabolomic layers from
IBDMDB to test whether the taxa identified as loop anchors also exhibit
coordinated functional disruption, providing the mechanistic link from
co-occurrence topology to butyrate synthesis and immune regulation.

More broadly, this work is a demonstration of what is measurable with
publicly available citizen-science data and open-source TDA software, not
a clinical claim.
The finding that \ibd-associated microbiomes are topologically simpler is
reproducible across cohorts and sequencing technologies, robust to
demographic confounding, grounded in an inflammation biomarker, and
interpretable at the level of named taxa.
It invites the question of whether topological complexity is a property
worth restoring therapeutically, and whether interventions that shift
microbiome composition (FMT, dietary modulation, precision probiotics) do
so by rebuilding co-occurrence loop structure or merely by altering
taxon abundances.
That question cannot be answered with cross-sectional data; it requires
the kind of dense longitudinal sampling that ongoing consortia are now
making feasible.
