%=============================================================================
% Topological Simplification in IBD-Associated Microbial Co-occurrence Networks
% American Gut Project (n = 3,409) — February 2026
%=============================================================================

\documentclass[11pt, a4paper]{article}

% ── Encoding & fonts ──────────────────────────────────────────────────────────
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{lmodern}
\usepackage{microtype}           % better typography, fewer overfull boxes

% ── Mathematics ───────────────────────────────────────────────────────────────
\usepackage{amsmath, amssymb, amsthm}

% ── Page geometry ─────────────────────────────────────────────────────────────
\usepackage[
  a4paper,
  top=25mm, bottom=28mm,
  left=28mm, right=28mm,
  headheight=14pt
]{geometry}

% ── Color ─────────────────────────────────────────────────────────────────────
\usepackage[dvipsnames]{xcolor}
\definecolor{darkblue}{RGB}{15, 52, 96}
\definecolor{midblue}{RGB}{52, 120, 180}
\definecolor{lightgray}{RGB}{245, 245, 248}
\definecolor{accentred}{RGB}{180, 50, 40}
\definecolor{linkblue}{RGB}{30, 100, 200}

% ── Hyperlinks ────────────────────────────────────────────────────────────────
\usepackage[
  colorlinks = true,
  linkcolor  = darkblue,
  citecolor  = midblue,
  urlcolor   = linkblue,
  pdftitle   = {Persistent Homology Reveals Topological Simplification in IBD-Associated Microbial Co-occurrence Networks},
  pdfauthor  = {Adam Levine},
  pdfsubject = {Microbiome Topology; TDA; IBD; American Gut Project},
  pdfkeywords= {persistent homology; microbiome; IBD; co-occurrence networks; topological data analysis},
  bookmarksnumbered = true
]{hyperref}

% ── Headers & footers ─────────────────────────────────────────────────────────
\usepackage{fancyhdr}
\pagestyle{fancy}
\fancyhf{}
\renewcommand{\headrulewidth}{0.4pt}
\renewcommand{\footrulewidth}{0.4pt}
\fancyhead[L]{\small\textcolor{darkblue}{\textit{Levine — Microbiome TDA}}}
\fancyhead[R]{\small\textcolor{darkblue}{\textit{February 2026 · Preprint}}}
\fancyfoot[C]{\small\textcolor{gray}{\thepage}}
\fancyfoot[L]{\small\textcolor{gray}{American Gut Project · AGP · $n = 3{,}409$}}
\fancyfoot[R]{\small\textcolor{gray}{\href{https://github.com/adam-jfkhs/Microbiome-TDA}{github.com/adam-jfkhs/Microbiome-TDA}}}

% Plain style for title page
\fancypagestyle{plain}{
  \fancyhf{}
  \renewcommand{\headrulewidth}{0pt}
  \renewcommand{\footrulewidth}{0.4pt}
  \fancyfoot[C]{\small\textcolor{gray}{\thepage}}
  \fancyfoot[R]{\small\textcolor{gray}{\href{https://github.com/adam-jfkhs/Microbiome-TDA}{github.com/adam-jfkhs/Microbiome-TDA}}}
}

% ── Figures & tables ──────────────────────────────────────────────────────────
\usepackage{graphicx}
\usepackage{booktabs}
\usepackage{array}
\usepackage{caption}
\usepackage{subcaption}
\usepackage{float}
\usepackage{multirow}

\captionsetup{
  labelfont = {bf, color=darkblue},
  textfont  = {small},
  margin    = 8pt,
  skip      = 6pt
}

% ── Line spacing & paragraph ──────────────────────────────────────────────────
\usepackage{setspace}
\setstretch{1.15}
\setlength{\parskip}{4pt}
\setlength{\parindent}{14pt}

% ── References ────────────────────────────────────────────────────────────────
\usepackage[numbers, sort&compress]{natbib}

% ── Misc ──────────────────────────────────────────────────────────────────────
\usepackage{enumitem}
\usepackage{lipsum}           % placeholder text — remove when writing is done

% ── Section styling ───────────────────────────────────────────────────────────
\usepackage{titlesec}
\titleformat{\section}
  {\large\bfseries\color{darkblue}}
  {\thesection}{1em}{}[\titlerule]
\titleformat{\subsection}
  {\normalsize\bfseries\color{midblue}}
  {\thesubsection}{1em}{}
\titleformat{\subsubsection}
  {\normalsize\itshape\color{darkblue}}
  {\thesubsubsection}{1em}{}

% ── Custom commands ───────────────────────────────────────────────────────────
\newcommand{\pval}[1]{\ensuremath{p = #1}}
\newcommand{\fdrpval}[1]{\ensuremath{p_{\text{FDR}} = #1}}
\newcommand{\cohend}[1]{\ensuremath{d = #1}}
\newcommand{\Hone}{\ensuremath{H_1}}
\newcommand{\Hzero}{\ensuremath{H_0}}
\newcommand{\agp}{\textsc{agp}}
\newcommand{\clr}{\textsc{clr}}
\newcommand{\vr}{\textsc{vr}}
\newcommand{\tda}{\textsc{tda}}
\newcommand{\ibd}{\textsc{ibd}}
\newcommand{\Ibd}{\textsc{Ibd}}
\newcommand{\abx}{\textsc{abx}}
\newcommand{\bh}{\textsc{bh}}
\newcommand{\todo}[1]{\textcolor{accentred}{\textbf{[TODO: #1]}}}

%=============================================================================
% TITLE BLOCK
%=============================================================================

\title{%
  \vspace{-10mm}
  {\LARGE\bfseries\color{darkblue}
    Persistent Homology Reveals Topological Simplification\\[4pt]
    in \textsc{ibd}-Associated Microbial Co-occurrence Networks}\\[8pt]
  {\large\normalfont\color{gray}
    A Paired Bootstrap Analysis of the American Gut Project ($n = 3{,}409$)\\
    with Covariate Matching and Taxa-Set Sensitivity Testing}
}

\author{%
  \textbf{Adam Levine}\\[3pt]
  \small\textit{Independent Researcher}\\[2pt]
  \small\href{mailto:adamwasnothere292@gmail.com}{adamwasnothere292@gmail.com}\\[4pt]
  \small\texttt{\href{https://github.com/adam-jfkhs/Microbiome-TDA}{github.com/adam-jfkhs/Microbiome-TDA}}
}

\date{%
  \small\textbf{Preprint} · February 2026\\[2pt]
  \small Not peer reviewed
}

%=============================================================================
\begin{document}
%=============================================================================

\maketitle
\thispagestyle{plain}

%─────────────────────────────────────────────────────────────────────────────
\begin{abstract}
\noindent\textbf{Background.}
Microbial co-occurrence networks encode higher-order interactions among gut
bacteria, yet conventional diversity metrics (Shannon entropy, Bray--Curtis
dissimilarity) are insensitive to the loop-level structure of these networks.
Persistent homology, a technique from topological data analysis (\tda),
extracts exactly this information: it tracks how many connected components,
loops (\Hone\ features), and voids form and dissolve across all correlation
thresholds simultaneously.

\noindent\textbf{Methods.}
We applied persistent homology to Spearman co-occurrence networks derived
from 3,409 stool samples from the American Gut Project (\agp), preprocessed
with centered log-ratio (\clr) transformation and filtered to a globally fixed
set of top-80 prevalent taxa.
We used a paired bootstrap design (200 iterations, $n = 100$ per group per
iteration) to obtain distributions of six scalar topological features per group.
Group comparisons were assessed via sign-flip label permutation
(500 shuffles; primary $p$-value) and Wilcoxon signed-rank test
(secondary/confirmatory), with Benjamini--Hochberg false discovery rate
correction applied across 18 tests (3 comparisons $\times$ 6 features) per
subset.
Confounding by age, sex, and body mass index was addressed by repeating all
analyses in matched subsamples.
Robustness was evaluated by repeating the full pipeline at taxon-set sizes
$N \in \{50, 80, 120\}$.

\noindent\textbf{Results.}
Inflammatory bowel disease (\ibd; $n = 165$ UC + Crohn's vs.\ $n = 3{,}084$
healthy) was associated with dramatically lower topological complexity across
all six \Hone\ features (Cohen's $d = 0.75$--$2.38$; $p_{\text{FDR}} <
0.005$ at all taxa sizes), a finding that was fully preserved after
age/sex/BMI matching.
Plant-based diet was associated with moderately longer-lived co-occurrence
loops (mean lifetime $d = 0.65$; $p_{\text{FDR}} = 0.004$), robust at
$N \in \{50, 80\}$.
Antibiotic recency showed limited signal (1--3 of 6 features, variable across
taxa sizes) that emerged only in matched analyses.

\noindent\textbf{Conclusions.}
\Ibd-associated gut microbiomes are topologically simpler than healthy
controls, suggesting that the disease state reshapes, rather than merely
reduces, the diversity of co-occurrence relationships.
This finding is robust to taxa-set size and demographic confounders.
Persistent homology provides a structurally informative complement to
existing microbiome diversity metrics.

\smallskip\noindent
\textbf{Keywords:} persistent homology, gut microbiome, inflammatory bowel
disease, co-occurrence networks, topological data analysis, American Gut
Project, bootstrap, FDR
\end{abstract}

\newpage
\tableofcontents
\newpage

%─────────────────────────────────────────────────────────────────────────────

\input{sections/01_introduction}
\input{sections/02_methods}
\input{sections/03_results}
\input{sections/04_discussion}
\input{sections/05_conclusion}

%─────────────────────────────────────────────────────────────────────────────
\newpage
\bibliographystyle{unsrtnat}
\bibliography{references}

%=============================================================================
\end{document}
%=============================================================================
